### 2025-03-15 paper/sections/abstract.qmd

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the stakeholder requirements. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

>>>>

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be (1) plausible with respect to the underlying data and (2) actionable with respect to the user-defined mutability constraints. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

### 2025-03-15 paper/sections/discussion.qmd

As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.  
`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.  
`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Nonetheless, these phenomena are not specific to CT.

>>>>

As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.  
`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature (see, e.g., [@venkatasubramanian2020philosophical]) impacting the capacity of CT to fulfill its intended goal.  
`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Nonetheless, these phenomena are not specific to CT.

### 2025-03-15 paper/sections/method.qmd

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE. One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.

@@

In all cases the counterfactuals are valid but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. In panel (c), the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

>>>>

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing that defines an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing that considers misclassification as the key distinguishing feature of AE. One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.

@@

In all cases the counterfactuals are all valid but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. In panel (c), the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.
